Staged typed type-preserving tagless interpreters, \emph{finally}

### 2007-03-24 matrix/tagless-final.tex
finished drafting Sec 2

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath}
\usepackage{natbib1}

@@

\begin{code}
module type Symantics = sig
  type ('c,'dv) repr
  val int  : int  -> ('c,int) repr
  val bool : bool -> ('c,bool) repr
  val add  : ('c,int) repr -> ('c,int) repr -> ('c,int) repr
  ...
  val if_ : ('c,bool) repr ->
             (unit -> ('c,'da) repr) ->
             (unit -> ('c,'da) repr) -> ('c,'da) repr

  val lam : (('c,'da) repr -> ('c,'db) repr) 
          -> ('c,'da->'db) repr
  val app : ('c,'da->'db) repr
    -> ('c,'da) repr -> ('c,'db) repr
  val fix : (('c,'da->'db) repr -> ('c,'da->'db) repr) 
            -> ('c,'da->'db) repr
end
\end{code}

Some differences are in the additional type parameter |c| (the
environment classifier, Walid03) to be used later when we get to code
generation (required by MetaOCaml, part of the type of the
code). MetaOCaml is a call-by-value language, and thus we have to eta-expand
the fix-point combinator and encapsulate the branches of the |if_|
into thunks to prevent their evaluation.

The power example and our sample term |test1| are encoded as
\begin{code}
module EX(S: Symantics) = struct
 open S
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = 
   lam (fun x ->fix (fun self -> lam (fun n ->
         if_ (leq n (int 0)) (fun () -> int 1)
             (fun () -> mul x (app self (add n (int (-1))))))))

 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end
\end{code}
The dummy unit argument in |testpowfix| is the artefact of MetaOCaml
(which we will be using to compile our tests in Sections below). To be
more precise, the environment qualifier (which appears as the |'c|
type parameter in the signature of Symantic functions) must remain
uninstantiated in order for us to `run' the code. That means, the type
must be polymorhic. The value restriction then dictates 
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

>>>>

\documentclass[preprint]{sigplanconf}
\usepackage{amsmath}
\usepackage{natbib1}
\usepackage{comment}

@@

\begin{code}
module type Symantics = sig
  type ('c,'dv) repr
  val int  : int  -> ('c,int) repr
  val bool: bool -> ('c,bool) repr
  val add : ('c,int) repr-> ('c,int) repr-> ('c,int) repr
  val mul : ('c,int) repr-> ('c,int) repr-> ('c,int) repr
  val leq : ('c,int) repr-> ('c,int) repr-> ('c,bool) repr
  val if_ : ('c,bool) repr ->
             (unit -> ('c,'da) repr) ->
             (unit -> ('c,'da) repr) -> ('c,'da) repr

  val lam : (('c,'da) repr -> ('c,'db) repr) 
          -> ('c,'da->'db) repr
  val app : ('c,'da->'db) repr
    -> ('c,'da) repr -> ('c,'db) repr
  val fix : (('c,'da->'db) repr -> ('c,'da->'db) repr) 
            -> ('c,'da->'db) repr
end;;
\end{code}

Some differences are in the additional type parameter |c| (the
environment classifier, Walid03) to be used later when we get to code
generation (required by MetaOCaml, part of the type of the
code). MetaOCaml is a call-by-value language, and thus we have to 
$\eta$-expand
the fix-point combinator and encapsulate the branches of the |if_|
into thunks to prevent their evaluation.

The power example and our sample term |test1| are encoded as
\begin{code}
module EX(S: Symantics) = struct
 open S
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = 
   lam (fun x ->fix (fun self -> lam (fun n ->
         if_ (leq n (int 0)) (fun () -> int 1)
             (fun () -> mul x (app self 
                                   (add n (int (-1))))))))
 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
The dummy unit argument in |testpowfix| is the artefact of MetaOCaml
(which we will be using to compile our tests in Sections below). To be
more precise, the environment qualifier (which appears as the |'c|
type parameter in the signature of Symantic functions) must remain
uninstantiated in order for us to `run' the code. That means, the type
must be polymorhic. The value restriction then dictates 
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
rank-2 record types of OCaml to maintain the necessary polymorphism.

@@

The first instatiation of the Symantics structure is the interpreter:
evaluating the object (DSL) term to its value in the meta-language. 
The following module essentially interprets (gives one-to-one
correspondence of) each operation of the object language as the
corresponding operation in the meta-language.
\begin{code}
module R  = struct
  type ('c,'dv) repr = 'dv    (* absolutely no wrappers *)
  let int (x:int) = x
  let bool (b:bool) = b
  let add e1 e2 = e1 + e2
  let mul e1 e2 = e1 * e2
  let leq x y = x <= y
  let if_ eb et ee = if eb then (et ()) else (ee ())

  let lam f = f
  let app e1 e2 = e1 e2
  let fix f = let rec self n = f self n in self
end;;
\end{code}

This interpreter is patently tagless: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. We can run our
examples by instatiating the |EX| functor with |R|.
\oleg{Jacques: you mentioned "Boxes go Bananas" by Washburn and
  Weirich. Do you want to comment more about that paper here?}

\begin{code}
  module EXR = EX(R);;
\end{code}

and so |EXR.test1 ()| evaluates to untagged boolean value |true|.
In Haskell, we define a |newtype R a = R a| and 
|instance Symantics R|
with the straightforward implementation. We must stress that |R| is
a |newtype|. Although it may look like a tag, it has no run-time
representation (erased during the compilation) and pattern-matching
against |R| cannot ever fail and so is operationally identity and
compiled that way.

Proposition: If |e| is the term in the meta-language has the type
|t|, evaluation of that term in the interpreter R will either continue
idenfinetely or termninate with a value of the same type |t|.

We'd like show a different interpreter: it interprets each terms of
the object language to its size (defined as the number of term
constructors). The following is a slightly abbreviated code (see the
accompanying source code for the complete definition)

\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int) = 1
  let mul e1 e2 = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1

  let lam f = (f 0) + 1
  let app e1 e2 = e1 + e2 + 1
  let fix f = (f 0) + 1
end;;
\end{code}

Now the expression
\begin{code}
  let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless, it is also
total. It `evaluates' even seemingly divergent terms like\\
|app (fix (fun self -> self)) (int 1)|.

\begin{comment}
module EX1(S: Symantics) = struct
 open S
 let tfix () = app (fix (fun self -> self)) (int 1)
end;;
let module E =EX1(R) in E.tfix ();;
let module E =EX1(L) in E.tfix ();;
\end{comment}
